---
title: "Ecological models in NIMBLE: Occupancy"
subtitle: "NIMBLE 2021 Virtual Workshop"
author: "NIMBLE Development Team"
date: "May 2021"
output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
library(nimble)
```

Occupancy models in `nimbleEcology`
=====

- `probOcc` = probability the site is occupied
- `probDetect[t]` = detection probability at time t.

### Variants

- `dOcc_s`: time-independent (scalar) detection probability
- `dOcc_v`: time-varying (vector) detection probability

### Example

```{r, eval=FALSE}
y[i, 1:T] ~ dOcc_v(probOcc = psi[i], 
                   probDetect = p[i, 1:T], len = T)
```

- `i` indexes site.

Dynamic occupancy models in `nimble`
=====

- `probPersist[t]` = persistence probability (occupied $\rightarrow$ occupied)
- `probColonize[t]` = colonization probability (unoccupied $\rightarrow$ occupied)
- `p[t, k]` = detection probability in year (primary) t, sampling visit k (secondary occasion)
- `start[t]` = first sampling occasion in year t.
- `end[t]` = last sampling occasion in year t.

### Variants

- `dDynOcc_[s|v][s|v][s|v|m]`
- first `[s|v]` is for persistence probability (time-independent or time-varying)
- second `[s|v]` is for colonization probability (time-independent or time-varying)
- third `[s|v|m]` is for detection probability (time-independent, time-varying, or time-visit-varying)

### Example

```{r, eval=FALSE}
y[1:num_sites, 1:num_years] ~ dDynOcc_svv(probPersist = psi, 
                                          probColonize = gamma[1:num_years],
                                          p = p[1:num_years])
```
