---
title: "Ecological models in NIMBLE: Hidden Markov models"
subtitle: "NIMBLE 2021 Virtual Workshop"
author: "NIMBLE Development Team"
date: "May 2021"
output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
has_nimbleEcology <- require(nimbleEcology)
has_compareMCMCs <- require(compareMCMCs)
if(!has_nimbleEcology)
  message("This module will use nimbleEcology, which you don't have installed.")
doDHMMexample <- TRUE
```


Hidden Markov models (Intro)
=====

Hidden Markov models (HMMs) are a general statistical framework for:

- systems with discrete latent states (usually) and discrete time steps

    * probabilities of transitioning between states at each time step

- discrete observation categories and imperfect observation (misclassification)
    
    *  probabilities of being observed in one category given the (true) latent state.

- If transition and/or observation probabilities are time-dependent, we call it a "Dynamic HMM" (DHMM).

Usually people use "Hidden Markov model" for discrete states and "state-space model" for the equivalent ideas with continuous states, but sometimes "Hidden Markov model" has been used for the latter.  They are also related to "partially observed Markov processes" and related terms.

Hidden Markov models (in capture-recapture)
=====

- latent states are often discrete spatial locations, breeding status, or other categories in combination with demographic categories such as juvenile, adult or dead and male or female.
- observation categories may include combinations of states (e.g. if juvenile and adult can't be distinguished in the field).
- HMMs encompass multi-state and multi-event capture-recapture.

Hidden Markov models (notation)
=====

- `T[i, j, t]`: probability of state change $i \rightarrow j$ in time step $t \rightarrow t+1$.
- `O[i, j, t]`: probability of observing state $i$ as category $j$ at time $t$.
- `init[i]`   : probability of being in state $i$ at $t = 1$.
- First observation comes from first state.

(In time-independent cases, the third index is dropped.)

**Orientation of rows vs. columns ("from state" and "to state") may differ from other examples you've seen.**

Orientation of "T[ from state, to state ]" is standard for HMMs but opposite of typical matrix population models.

Hidden Markov models (variants in `nimble`)
=====

d[D]HMM[o]: 

- "D" indicates time-varying transition probabilities.
- "o" indicates time-varying observation probabilities.

- dHMM: Neither is time-varying
- dHMMo: Only observations are time-varying
- dDHMM: Only transitions are time-varying
- dDHMMo: Both are time-varying

Hidden Markov models (typical implementation)
=====

A typical implementation in model code would be like:
```{r, eval=FALSE}
## Not indexing transition and observation probabilities by individual, for simplicity of example
for(t in 1:T) {
  for(i in 1:num_states) {
    for(j in 1:num_states) {
      T[i, j, t] <- some_calculation() # Involving survival and transition probabilities 
                                       # that may depend on explanatory variables or random effects
    }
  }
}
## Some similar definition of O, observation probabilities
for(k in 1:num_individuals) {
  for(t in 2:T) {
    z[k, t] ~ dcat(T[z[t-1], 1:num_states, t])
  }
  for(t in 1:T) {
    y[k, t] ~ dcat(O[z[t], 1:num_states, t])
  }
}
```

Let's discuss:

- Computational cost of MCMC sampling for `z[t]` based on `dcat`.
- Dependencies of survival and transition probabilities (or their underlying parameters).

Hidden Markov models (worked example)
=====

### Orchids (ladyslipper) example

- From [*Bayesian Population Analysis Using WinBUGS*](https://www.vogelwarte.ch/de/projekte/publikationen/bpa/) (K&#233;ry and Schaub 2011), aka BPA.
- Orchids modeled in states "vegetative", "flowering", "dormant", "dead"
- Observations: "seen vegetative", "seen flowering", "not seen".
- k = 11 annual observation periods 
- n = 250 unique plants
- n = 236 plants seen before the last observation time

This is a multi-state model because the state when not seen is unknown.  The following would also work for multi-event models.

Obviously multi-state models can be more complicated than this.


Setup data
=====
Data and code (possibly adapted) from authors.
```{r}
CH <- as.matrix(read.table("orchids.txt", sep=" ", header = FALSE))
head(CH) ## Glimpse of data
n_occasions <- dim(CH)[2]
## Compute vector with occasion of first capture
f <- numeric()
for (i in 1:dim(CH)[1])
  f[i] <- min(which(CH[i,]!=0))

## Modification for NIMBLE:
CH <- CH[which(f!=11), ]  ## remove all individuals not seen until the last occasion: They contribute no information and create problems for the dDHMM version.
## Reset f from the reduced CH
f <- numeric()
for (i in 1:dim(CH)[1])
  f[i] <- min(which(CH[i,]!=0))

## Recode CH matrix: note, a 0 is not allowed by WinBUGS!
## 1 = seen vegetative, 2 = seen flowering, 3 = not seen 
rCH <- CH  # Recoded CH 
rCH[rCH==0] <- 3 

## Function to create known latent states z 
known_state_ms <- function(ms, notseen){
    ## notseen: label for not seen 
    state <- ms 
    state[state==notseen] <- NA 
    for (i in 1:dim(ms)[1]){
        m <- min(which(!is.na(state[i,]))) 
        state[i,m] <- NA 
    }
    return(state) 
}
## Bundle data 
orchids_data <- list(y = rCH,
                  f = f,
                  n_occasions = dim(rCH)[2],
                  nind = dim(rCH)[1],
                  z = known_state_ms(rCH, 3)) 

ms_init_z <- function(ch, f){
   for (i in 1:dim(ch)[1]){ch[i,1:f[i]] <- NA}
   states <- max(ch, na.rm = TRUE)
   known.states <- 1:(states-1)
   v <- which(ch==states)
   ch[-v] <- NA
   ch[v] <- sample(known.states, length(v), replace = TRUE)
   return(ch)
   }

## Initial values 
orchids_inits <- function(){
    list(s = runif((dim(rCH)[2]-1), 0, 1),
         z = ms_init_z(rCH, f)
         )}

```

Original model code
=====

(Hint: Look for wasted computation, but also remember the authors intended this as a teaching example so they must have wanted it to be general.)

- `ps[1, i, t, 2]` = probability of transition from state 1 to 2 for individual `i` from time `t` to time `t+1`.
- `po[1, i, t, 2]` = probability of being observed in state 2 when in state 1 for individual `i` at time `t`.
- `s[t]` = survival probability at time `t`.
- `psiV[j]` = probability of transition from state 1 (Vegetative) to state j.
- `psiF[j]` = probability of transition from state 2 (Flowering) to state j.
- `psiD[j]` = probability of transition from state 3 (Dormant) to state j.

```{r}
orchids_code <- nimbleCode({
  ## -------------------------------------------------
  ## Parameters:
  ## s: survival probability 
  ## psiV: transitions from vegetative 
  ## psiF: transitions from flowering 
  ## psiD: transitions from dormant 
  ## -------------------------------------------------
  ## States (S):
  ## 1 vegetative 
  ## 2 flowering 
  ## 3 dormant 
  ## 4 dead 
  ## Observations (O):  
  ## 1 seen vegetative 
  ## 2 seen flowering 
  ## 3 not seen 
  ## -------------------------------------------------
  ## Priors and constraints 
  ## Survival: uniform 
  for (t in 1:(n_occasions-1)){  
    s[t] ~ dunif(0, 1) 
  }
  ## Transitions: gamma priors 
  for (i in 1:3){
    a[i] ~ dgamma(1, 1) 
    psiD[i] <- a[i]/sum(a[1:3]) 
    b[i] ~ dgamma(1, 1) 
    psiV[i] <- b[i]/sum(b[1:3]) 
    c[i] ~ dgamma(1, 1) 
    psiF[i] <- c[i]/sum(c[1:3]) 
  }
  ## Define state-transition and observation matrices 	
  for (i in 1:nind){
    ## Define probabilities of state S(t+1) given S(t) 
    for (t in 1:(n_occasions-1)){
      ps[1,i,t,1] <- s[t] * psiV[1]
      ps[1,i,t,2] <- s[t] * psiV[2]
      ps[1,i,t,3] <- s[t] * psiV[3]
      ps[1,i,t,4] <- 1-s[t]
      ps[2,i,t,1] <- s[t] * psiF[1]
      ps[2,i,t,2] <- s[t] * psiF[2]
      ps[2,i,t,3] <- s[t] * psiF[3]
      ps[2,i,t,4] <- 1-s[t]
      ps[3,i,t,1] <- s[t] * psiD[1]
      ps[3,i,t,2] <- s[t] * psiD[2]
      ps[3,i,t,3] <- s[t] * psiD[3]
      ps[3,i,t,4] <- 1-s[t]
      ps[4,i,t,1] <- 0 
      ps[4,i,t,2] <- 0 
      ps[4,i,t,3] <- 0 
      ps[4,i,t,4] <- 1 
      ## Define probabilities of O(t) given S(t) 
      po[1,i,t,1] <- 1 
      po[1,i,t,2] <- 0 
      po[1,i,t,3] <- 0 
      po[2,i,t,1] <- 0 
      po[2,i,t,2] <- 1 
      po[2,i,t,3] <- 0 
      po[3,i,t,1] <- 0 
      po[3,i,t,2] <- 0 
      po[3,i,t,3] <- 1 
      po[4,i,t,1] <- 0 
      po[4,i,t,2] <- 0 
      po[4,i,t,3] <- 1 
    } #t 
  } #i 
  ## Likelihood
  for (i in 1:nind){
    ## Define latent state at first capture 
    z[i,f[i]] <- y[i,f[i]]
    for (t in (f[i]+1):n_occasions){
      ## State process: draw S(t) given S(t-1)
      z[i,t] ~ dcat(ps[z[i,t-1], i, t-1, 1:4]) 
      ## Observation process: draw O(t) given S(t) 
      y[i,t] ~ dcat(po[z[i,t], i, t-1, 1:3]) 
    } #t 
  } #i
})
set.seed(123)
orchids_info <- list(code=orchids_code, constants=orchids_data, inits=orchids_inits())
```

Run original version in `nimble` and `jags`
=====

Some of the runs below may generate model initialization warnings.  They shouldn't be a problem.

```{r, eval = doDHMMexample}
orchids_result <- compareMCMCs(
  orchids_info,
  MCMCs = c("nimble", "jags"),
  monitors = c("s", "psiV", "psiF", "psiD"),
  MCMCcontrol = list(niter = 20000,
                     burnin = 2000)
)
orchids_result <- renameMCMC(orchids_result, "nimble_orig", "nimble")
orchids_result <- renameMCMC(orchids_result, "jags_orig", "jags")
```

Version without unnecessary (computationally wasteful) indexing over individuals and/or time
=====
```{r}
orchids_code_faster <- nimbleCode({
  ## -------------------------------------------------
  ## Parameters:
  ## s: survival probability 
  ## psiV: transitions from vegetative 
  ## psiF: transitions from flowering 
  ## psiD: transitions from dormant 
  ## -------------------------------------------------
  ## States (S):
  ## 1 vegetative 
  ## 2 flowering 
  ## 3 dormant 
  ## 4 dead 
  ## Observations (O):  
  ## 1 seen vegetative 
  ## 2 seen flowering 
  ## 3 not seen 
  ## -------------------------------------------------
  ## Priors and constraints 
  ## Survival: uniform 
  for (t in 1:(n_occasions-1)){  
    s[t] ~ dunif(0, 1) 
  }
  ## Transitions: gamma priors 
  for (i in 1:3){
    a[i] ~ dgamma(1, 1) 
    psiD[i] <- a[i]/sum(a[1:3]) 
    b[i] ~ dgamma(1, 1) 
    psiV[i] <- b[i]/sum(b[1:3]) 
    c[i] ~ dgamma(1, 1) 
    psiF[i] <- c[i]/sum(c[1:3]) 
  }
  ## Define state-transition and observation matrices 	
  ## Define probabilities of state S(t+1) given S(t) 
  for (t in 1:(n_occasions-1)){
    ps[1,t,1] <- s[t] * psiV[1]
    ps[1,t,2] <- s[t] * psiV[2]
    ps[1,t,3] <- s[t] * psiV[3]
    ps[1,t,4] <- 1-s[t]
    ps[2,t,1] <- s[t] * psiF[1]
    ps[2,t,2] <- s[t] * psiF[2]
    ps[2,t,3] <- s[t] * psiF[3]
    ps[2,t,4] <- 1-s[t]
    ps[3,t,1] <- s[t] * psiD[1]
    ps[3,t,2] <- s[t] * psiD[2]
    ps[3,t,3] <- s[t] * psiD[3]
    ps[3,t,4] <- 1-s[t]
    ps[4,t,1] <- 0 
    ps[4,t,2] <- 0 
    ps[4,t,3] <- 0 
    ps[4,t,4] <- 1 
  }
  ## Define probabilities of O(t) given S(t) 
  po[1,1] <- 1 
  po[1,2] <- 0 
  po[1,3] <- 0 
  po[2,1] <- 0 
  po[2,2] <- 1 
  po[2,3] <- 0 
  po[3,1] <- 0 
  po[3,2] <- 0 
  po[3,3] <- 1 
  po[4,1] <- 0 
  po[4,2] <- 0 
  po[4,3] <- 1 
  ## Likelihood
  for (i in 1:nind){
    ## Define latent state at first capture 
    z[i,f[i]] <- y[i,f[i]]
    for (t in (f[i]+1):n_occasions){
      ## State process: draw S(t) given S(t-1)
      z[i,t] ~ dcat(ps[z[i,t-1], t-1, 1:4]) 
      ## Observation process: draw O(t) given S(t) 
      y[i,t] ~ dcat(po[z[i,t], 1:3]) 
    } #t 
  } #i
})
```

```{r, eval=doDHMMexample}
set.seed(123)
orchids_info_faster <- list(code=orchids_code_faster,
                            constants=orchids_data, 
                            inits=orchids_inits())
```

Run version without unnecessary indexing (faster)
=====

Note: Some elements of `z` (with perfect observation categories) are provided as `data`.  Others as `inits`.

```{r, eval=doDHMMexample}
orchids_result_faster <- compareMCMCs(
  orchids_info_faster,
  MCMCs = c("nimble", "jags"),
  monitors = c("s", "psiV", "psiF", "psiD"),
  MCMCcontrol = list(niter = 20000,
                     burnin = 2000)
)
orchids_result_faster <- renameMCMC(orchids_result_faster, "nimble_faster", "nimble")
orchids_result_faster <- renameMCMC(orchids_result_faster, "jags_faster", "jags")
```

DHMM version
=====
```{r}
orchids_code_DHMM <- quote({
  ## -------------------------------------------------
  ## Parameters:
  ## s: survival probability 
  ## psiV: transitions from vegetative 
  ## psiF: transitions from flowering 
  ## psiD: transitions from dormant 
  ## -------------------------------------------------
  ## States (S):
  ## 1 vegetative 
  ## 2 flowering 
  ## 3 dormant 
  ## 4 dead 
  ## Observations (O):  
  ## 1 seen vegetative 
  ## 2 seen flowering 
  ## 3 not seen 
  ## -------------------------------------------------
  ## Priors and constraints 
  ## Survival: uniform 
  for (t in 1:(k-1)){  
    s[t] ~ dunif(0, 1)
  }
  ## Transitions: gamma priors 
  for (i in 1:3){
    a[i] ~ dgamma(1, 1) 
    psiD[i] <- a[i]/sum(a[1:3]) 
    b[i] ~ dgamma(1, 1) 
    psiV[i] <- b[i]/sum(b[1:3]) 
    c[i] ~ dgamma(1, 1) 
    psiF[i] <- c[i]/sum(c[1:3]) 
  }
  ## Define state-transition and observation matrices 	
  for (t in 1:(k-1)) {
    T[1,1,t] <- s[t] * psiV[1]
    T[1,2,t] <- s[t] * psiV[2]
    T[1,3,t] <- s[t] * psiV[3]
    T[1,4,t] <- 1-s[t]
    T[2,1,t] <- s[t] * psiF[1]
    T[2,2,t] <- s[t] * psiF[2]
    T[2,3,t] <- s[t] * psiF[3]
    T[2,4,t] <- 1-s[t]
    T[3,1,t] <- s[t] * psiD[1]
    T[3,2,t] <- s[t] * psiD[2]
    T[3,3,t] <- s[t] * psiD[3]
    T[3,4,t] <- 1-s[t]
    T[4,1,t] <- 0
    T[4,2,t] <- 0
    T[4,3,t] <- 0
    T[4,4,t] <- 1
  }
  O[1,1] <- 1 
  O[1,2] <- 0 
  O[1,3] <- 0 
  O[2,1] <- 0 
  O[2,2] <- 1 
  O[2,3] <- 0 
  O[3,1] <- 0 
  O[3,2] <- 0 
  O[3,3] <- 1 
  O[4,1] <- 0 
  O[4,2] <- 0 
  O[4,3] <- 1 
  for(i in 1:nind) {
    for(j in 1:4)
      init[i, j] <- y2[i, f[i]]==j ## y2 is the same as y, to avoid a cycle in graph (i.e. y[i, f[i]] depends on init[i, 1:3] which depends on y[i, f[i]])
  }
  for (i in 1:nind) {
    y[i, f[i]:k] ~ dDHMM(init = init[i, 1:4],
                         probObs = O[1:4,1:3],
                         probTrans = T[1:4,1:4,f[i]:(k-1)],  ## Note: possible length 1 issue
                         len = k-f[i]+1,
                         checkRowSums = 0)
  }
})
orchids_constants_DHMM <- list(f=f,
                               k=dim(rCH)[2],
                               nind=dim(rCH)[1])
orchids_data_DHMM <- list(y = rCH, y2 = rCH)
orchids_inits_DHMM <- function() {
  list(s = runif((dim(rCH)[2]-1), 0, 1),
       a = rep(1,3),
       b = rep(1,3),
       c = rep(1,3))
}
set.seed(123)
orchids_info_DHMM <- list(code=orchids_code_DHMM, 
                          constants=orchids_constants_DHMM,
                          data = orchids_data_DHMM,
                          inits=orchids_inits_DHMM())
```

```{r, eval=doDHMMexample}
orchids_result_DHMM <- compareMCMCs(
  orchids_info_DHMM,
  MCMCs = c("nimble"),
  monitors = c("s", "psiV", "psiF", "psiD"),
  MCMCcontrol = list(niter = 20000,
                     burnin = 2000)
)
orchids_result_DHMM <- renameMCMC(orchids_result_DHMM, "nimble_DHMM", "nimble")
```

Make results figure
=====
```{r, eval=doDHMMexample, echo=FALSE}
# Code for generating results that come with slides
make_MCMC_comparison_pages(c(orchids_result, orchids_result_faster, orchids_result_DHMM),
                           dir = "orchid_results_with_slides",
                           modelName = "orchids")
```

```{r, eval=FALSE}
# Code for your own results:
make_MCMC_comparison_pages(c(orchids_result, orchids_result_faster, orchids_result_DHMM),
                           dir = "your_orchid_results",
                           modelName = "orchids")
```

Results that come with these slides are [here](orchid_results_with_slides/orchids.html).

Results if you generate them yourself will be [here](your_orchid_results/orchids.html).
