---
title: "Introduction to programming with nimbleFunctions"
subtitle: "NIMBLE 2021 Virtual Workshop"
author: "NIMBLE Development Team"
date: "May 2021"

output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nimble)
```

# Introduction 

*nimbleFunctions* are at the heart of NIMBLE. They are the way that algorithms are implemented. They can also be used for

 - user-defined BUGS distributions (already seen),
 - user-defined BUGS functions (already seen)
 - user-defined MCMC samplers (Module 9b), and
 - compiling parts of R (not shown), without reference to a model.

But their main purpose is providing a way for developers to implement algorithms.

# Two-stage evaluation in nimbleFunctions


Say we want a nimbleFunction to calculate some nodes and their dependencies.

```{r}
calcDeps <- nimbleFunction(
  setup = function(model, nodes) { # setup function gives first stage of evalution
    calcNodes <- model$getDependencies(nodes)
  },
  run = function() {               # run function (or other methods) give sectond stage of evaluation
    ans <- model$calculate(calcNodes)
    return(ans)
    returnType(double())
  }
)
```

# Using the nimbleFunction

Let's try this on a very basic linear regression:

```{r}
regrCode <- nimbleCode({
  b0 ~ dnorm(0, sd = 100)
  b1 ~ dnorm(0, sd = 100)
  sigma ~ dunif(0, 100)
  for(i in 1:n)
     y[i] ~ dnorm(b0 + b1*x[i], sd = sigma)
})

set.seed(1)
n <- 100
x <- runif(n)
b0_true <- 0.3
b1_true <- 0.5
sigma_true <- 0.25
y <- rnorm(n, b0_true + b1_true*x, sigma_true)

regrModel <- nimbleModel(regrCode, data = list(y = y), constants = list(n = n, x = x),
                     inits = list(b0 = 1, b1 = 0, sigma = 1))
```

```{r}
calcDeps_regr <- calcDeps(regrModel, c('b0', 'b1', 'sigma'))
calcDeps_regr$run()   ## second stage: run code
```

# Some notes

We could compiled `calcDeps_regr`.  Then `run` would become a method of a C++ class.

Notice that:

- `calcDeps_regr` is an R reference class object of a custom-generated class.
- `model` and `calcNodes` are fields in the class

```{r}
class(calcDeps_regr) ## We could have used nimbleFunction's name argument to set the class name
calcDeps_regr$calcNodes[1:20]
```

# Demonstration of uncompiled browsing/debugging.


```{r, eval=FALSE}
calcDeps <- nimbleFunction(
  setup = function(model, nodes) {
    browser()
    calcNodes <- model$getDependencies(nodes)
  },
  run = function() {
    browser()
    ans <- model$calculate(calcNodes)
    return(ans)
    returnType(double())
  }
) ## warning about not being able to compiled with browser() is expected.
```

Stepping through debugging from `browser()` will not work well in Rmarkdown, so this code is not evaluated.  Run it in your own R session.
```{r, eval=FALSE}
calcDeps_regr <- calcDeps(model, c('b0', 'b1', 'sigma')) ## We'll see the setup code followed by internal code.
calcDeps_regr$run()
```

# More about nimbleFunctions

- Without setup code, a `nimbleFunction` becomes an R function (uncompiled) and a C++ function (compiled).
- With setup code, a `nimbleFunction` becomes an R reference class definition (uncompiled) and a C++ class definition (compiled).
    - `nimbleFunction` returns a generator (aka constructor, aka initializer) of new class objects.

nimbleFunction class definitions (i.e with setup code):

- `setup` is always executed in R.
    - Typically one-time, high-level processing such as querying model structure.
- `run` and other methods can be run uncompiled (in R) or compiled (via C++).
    - Typically repeated "actual algorithm" calculations such as MCMC sampler updates.
    - Can operate models.
- Any objects (e.g. `calcNodes` and `model`) in `setup` can be used in `run`.
    - Internally, these are automatically set up as class member data.
    - You do not need to explicitly declare class member data.
    - Nodes used in model operations are "baked in" (aka partially evaluated) during compilation. 
        - Node vectors must be created in setup code and used in run code.
        - They can't be dynamically modified in run code.


# Example: using nimbleFunctions for maximization

Suppose we wanted to maximize the likelihood of our regression model.

Let's consider how we would optimize the parameters in a model using a nimbleFunction. Basically, we'll just construct an objective function that we can then pass to R's *optim* function to do the actual numerical optimization. (NIMBLE also has an `optim()` that you can use within a nimbleFunction.)

Plan:

 - Setup code determines what needs to be calculated
 - Run code:
    - checks for correct parameter input length
    - puts parameter values into model
    - calculates the likelihood

# A nimbleFunction for the regression model likelihood

```{r, regr-objective, eval}
objective <- nimbleFunction(
    setup = function(model, nodes) {
        calcNodes <- model$getDependencies(nodes)
        elements <- model$expandNodeNames(nodes, returnScalarComponents = TRUE)
        n <- length(elements)
    },
    run = function(par = double(1)) {
        returnType(double(0))
        if(length(par) != n)
           stop("Input length does not match number of parameter elements.")
        values(model, nodes) <<- par   # assignment into non-local (nf) variables 
        ans <- model$calculate(calcNodes)  # local assignment
        return(ans)
    }
)
```

As discussed, this is actually a nimbleFunction *generator* -- we can't run it yet -- we need to create a specialized instance of the nimbleFunction that is tailored for some model, in our case the marginalized litters model. 

One caveat: we haven't dealt carefully with the constraint that the standard deviation must be positive.

# Specializing the nimbleFunction to the model

```{r, regr-specialized}
rObjective <- objective(regrModel, c('b0', 'b1', 'sigma'))
cRegrModel <- compileNimble(regrModel)   # remember to compile model first
cObjective <- compileNimble(rObjective, project = regrModel)
```

Now let's try using it.

```{r, regr-optimize}
system.time(optR <- optim(c(0, 0, 1), rObjective$run, control = list(fnscale = -1)))
system.time(optC <- optim(c(0, 0, 1), cObjective$run, control = list(fnscale = -1)))
optR
optC
```
