---
title: "nimbleFunction programming"
subtitle: "NIMBLE 2021 Virtual Workshop"
author: "NIMBLE Development Team"
date: "May 2021"

output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nimble)
if(!require(coda))
  warning("This Rmd file needs package coda")
has_compareMCMCs <- require(compareMCMCs)
if(!has_compareMCMCs)
  warning("This Rmd file uses package compareMCMCs from github.  Sections using it will be skipped.")
generate_orig_results <- FALSE
```

# Agenda

1. Review of nimbleFunctions with setupCode
2. nimbleFunctions for new (i.e., user-defined) samplers
   1. Design of NIMBLE's MCMC system
   2. Setting up a new sampler: A simple Metropolis-Hastings example
   3. Worked example: sampler that jointly updates intercepts and random effects in the E. cervi example.
3. modelValues for storing values from a model
4. nimbleFunctions for post-processing
   1. Worked example: posterior predictive checks
   2. Worked example: sampling from marginalized nodes
5. Calling out to R and C++ from nimbleFunctions


# Load Deer E. cervi example

```{r load-DeerEcervi}
source(file.path("..", "examples", "DeerEcervi", "load_DeerEcervi.R"), 
       chdir = TRUE)
set.seed(123)
DEmodel <- nimbleModel(DEcode,
                        constants = DEconstants,
                        data = list(Ecervi_01 = DeerEcervi$Ecervi_01),
                        inits = DEinits())
```

# Design of `nimble`'s MCMC system


Here is a figure of MCMC configuration and MCMCs:[nimble_MCMC_design.pdf](nimble_MCMC_design.pdf)

1. MCMC configuration object: Contains a list of sampler assignments, not actual samplers.
2. MCMC object: Contains a list of sampler objects.

To write new samplers, we need to understand:

- Two-stage evaluation of nimbleFunctions with setup code.
- Setup (configuration) rules for using a new MCMC sampler
- Run-time rules for management of model calculations and saved states.
- More about nimbleFunction programming.

# Example of MCMC configuration object

Look at MCMC configuration object for just a few nodes
```{r}
mcmcConf <- configureMCMC(DEmodel, nodes = 'farm_effect[1:3]')
class(mcmcConf)
ls(mcmcConf)
fe1_sampler <- mcmcConf$getSamplers("farm_effect[1]")[[1]]
class(fe1_sampler)
ls(fe1_sampler)
fe1_sampler$target
fe1_sampler$control
```

# What happens from an MCMC configuration object?


Eventually, an MCMC sampler object is created by a call like this (for adaptive random-walk MH):
```{r, eval=FALSE}
sampler_RW(model, mvSaved, target, control)
```

- This is stage one of two-stage evaluation.  It instantiates an object of a `sampler_RW` class.
- `model` is the model
- `mvSaved` is a `modelValues` object for keeping a set of saved model states. (more later)
- `target` is a vector of target node names
- `control` is a list of whatever sampler-specific configuration settings are needed.


# More about nimbleFunctions

- Without setup code, a `nimbleFunction` becomes an R function (uncompiled) and a C++ function (compiled).
- With setup code, a `nimbleFunction` becomes an R reference class definition (uncompiled) and a C++ class definition (compiled).
    - `nimbleFunction` returns a generator (aka constructor, aka initializer) of new class objects.

### nimbleFunction class definitions (i.e., with setup code)

- `setup` is always executed in R.
    - Typically one-time, high-level processing such as querying model structure.
- `run` and other methods can be run uncompiled (in R) or compiled (via C++).
    - Typically repeated "actual algorithm" calculations such as MCMC sampler updates.
    - Can operate models.
- Any objects (e.g. `calcNodes` and `model`) in `setup` can be used in `run`.
    - Internally, these are automatically set up as class member data.
    - You do not need to explicitly declare class member data.
    - Nodes used in model operations are "baked in" (aka partially evaluated) during compilation. 
        - Node vectors must be created in setup code and used in run code.
        - They can't be dynamically modified in run code.

# A basic Random-Walk Metropolis-Hastings sampler

```{r}
ourMH <- nimbleFunction(
  name = 'ourMH',                              # Convenient for class name of R reference class and generated C++ class
  contains = sampler_BASE,                     # There is a simple class inheritance system.
  setup = function(model, mvSaved, target, control) {                 # REQUIRED setup arguments
    scale <- if(!is.null(control$scale)) control$scale else 1         # Typical extraction of control choices
    calcNodes <- model$getDependencies(target)                        # Typical query of model structure
  },                                                                  # setup can't return anything
  run = function() {
    currentValue <- model[[target]]                                   # extract current value
    currentLogProb <- model$getLogProb(calcNodes)                     # get log "denominator" from cached values
    proposalValue <- rnorm(1, mean = currentValue, sd = scale)        # generate proposal value
    model[[target]] <<- proposalValue                                 # put proposal value in model
    proposalLogProb <- model$calculate(calcNodes)                     # calculate log "numerator" 
    logAcceptanceRatio <- currentLogProb - proposalLogProb            # log acceptance ratio
    # Alternative:
    # logAcceptanceRatio <- model$calculateDiff(calcNodes)
    accept <- decide(logMHR)                                          # utility function to generate accept/reject decision
    if(accept)                                                        # accept synchronize model -> mvSaved
      copy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
    else                                                              # reject: synchronize mvSaved -> model
      copy(from = mvSaved, to = model, row = 1, nodes = calcNodes, logProb = TRUE)
  },
  methods = list(                              # required method for sampler_BASE base class
    reset = function() {}
  )
)
```

# Rules for each sampler

### setup function
  
- The four arguments, named exactly as shown, are required.  This allows `buildMCMC` to create any sampler correctly.

### run function

- The `mvSaved` ("modelValues saved") has a saved copy of all model variables and log probabilities
- Upon entry, `run()` can assume:
    - the model is fully calculated (so `getLogProb` and `calculateDiff` make sense).
    - `mvSaved` and the model are synchronized (have the same values).
- Upon exit, `run()` must ensure those conditions are met.
    - That way the next sampler can operator correctly.
- Between entry and exit, `run()` can manipulate the model in any way necessary.

### reset function

- To match the `sampler_BASE` definition, all samplers must have a `reset()` function.


# Stepping through uncompiled execution


Version with `browser()`s

```{r}
ourMH_debug <- nimbleFunction(
  name = 'ourMH_debug',
  contains = sampler_BASE,
  setup = function(model, mvSaved, target, control) {
    browser()
    scale <- if(!is.null(control$scale)) control$scale else 1 
    calcNodes <- model$getDependencies(target)                
  },                                                          
  run = function() {
    browser()
    currentValue <- model[[target]]                           
    currentLogProb <- model$getLogProb(calcNodes)             
    proposalValue <- rnorm(1, mean = currentValue, sd = scale)
    model[[target]] <<- proposalValue                             
    proposalLogProb <- model$calculate(calcNodes)             
    logAcceptanceRatio <- currentLogProb - proposalLogProb    
    accept <- decide(logMHR)                                     
    if(accept)                                                   
      copy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
    else                                                         
      copy(from = mvSaved, to = model, row = 1, nodes = calcNodes, logProb = TRUE)
  },
  methods = list(                              
    reset = function() {}
  )
)
```

# Stepping through uncompiled execution

```{r}
mcmcConf <- configureMCMC(DEmodel, nodes = NULL) ## Make an empty configuration
mcmcConf$addSampler(type = "ourMH_debug", target = 'farm_effect[1]')
```

```{r, eval=FALSE}
# run this on your own to step through in debug (browser) mode.
mcmc <- buildMCMC(mcmcConf)
mcmc$run(5)
```

# Let's modify the basic RW sampler

A user wrote to nimble-users recently asking how to do MCMC sampling on subsets of a multivariate node, e.g., for missing data applications.

Unfortunately, `nimble` is not set up to do this. The user took our basic RW sampler and modified it for his need.

Let's open up [`MCMC_samplers.R`](https://github.com/nimble-dev/nimble/blob/devel/packages/nimble/R/MCMC_samplers.R) (and search for `sampler_RW`), so we can have the basic sampler in front of us.

Let's suppose we have this basic model with a node, say `y[1:3]` and `y[2]` is missing.

`nimble` easily handles nodes that are missing (assigning an MCMC sampler to them) but not multivariate nodes that are a mix of missing and non-missing. But we can work around that. 

```{r}
code <- nimbleCode({
     y[1:n] ~ dmnorm(mu[1:n], cov = Sigma[1:n, 1:n])
})
n <- 3
mu <- rep(0, n)
Sigma <- matrix(c(1, .9, -.7, .9, 1, -.6, -.7, -.6, 1), 3, 3)
Sigma

y <- c(0.2, NA, -0.5)
yFull <- y
yFull[2] <- 0
model <- nimbleModel(code, inits = list(y = yFull),
  constants = list(n = n, mu = mu, Sigma = Sigma))
conf <- configureMCMC(model, nodes = NULL)
# conf$addSampler(target = '???', type = 'conditional_RW', control = list(???))
```

- What information do we need from the user to set up the sampler?
- What do we need to do in the run code to sample `y[2]`?
   - Note: when we call `calculate`, we need to calculate the multivariate density of `y[1:5]`

# A sampler for an element of a multivariate node

Let's see [the sampler the user came up with](conditional_RW.R).

The key pieces are:

```
#### from setup code:

        ## set incorrect defaults so that it throws an error if not explicitly set
        index               <- extractControlElement(control, 'index',               1:2)

        ## checks on index
        if(length(index) != 1) stop('length of index must be 1')
        if(index < 1 | index > d) stop('index must be within number of dimensions of target')

#### from run code:

        currentValue <- model[[target]]
        propValue <- model[[target]]
        propLogScale <- 0
        if(logScale) { 
            propLogScale <- rnorm(1, mean = 0, sd = scale)
            propValue[index] <- currentValue[index] * exp(propLogScale)
        } else {
            propValue[index] <- rnorm(1, mean = currentValue[index],  sd = scale)
        }
        model[[target]] <<- propValue
```

# Using the new sampler

Using the new sampler in `nimble` is easy.

```{r, fig.width=8, fig.height=4, fig.cap=''}
source('conditional_RW.R')
index <- which(is.na(y))
conf$addSampler(target = 'y[1:3]', type = 'conditional_RW', control = list(index = index))
mcmc <- buildMCMC(conf)

cModel <- compileNimble(model)
cmcmc <- compileNimble(mcmc, project = model)

samples <- runMCMC(cmcmc, niter = 1000)
head(samples)
par(mfrow = c(1,2))
ts.plot(samples[ , index] )
hist(samples[ , index])
```

# Let's make an interesting sampler (1)

Let's work on a more involved case where we can't make such simple modifications to an existing sampler.

We'll put the pieces together for the E. cervi example as follows:

- Use the original parameterization.
- Write a sampler that proposes to add $\delta \sim N(0, \mbox{scale})$ to the two sex-specific intercepts and to subtract the same $\delta$ from every `farm_effect[i]` ($i = 1 \ldots 24$).
- This is a scalar sampler in rotated coordinates.
- The coordinate transformation is linear, so there is no determinant of a Jacobian matrix to incorporate.  In general one needs to be careful to use distribution theory correctly if non-linear coordinate transformations are involved.  We are covering the software, not the math.
- We want proposal scale to be adaptive (self-tuning as the MCMC proceeds).  We will just copy from NIMBLE's `sampler_RW` to implement that.


# Let's make an interesting sampler (2)


```{r}
ourSampler <- nimbleFunction(
  name = 'ourSampler',
  contains = sampler_BASE,
  setup = function(model, mvSaved, target, control) {
    ## control list extraction
    offsetNodes <- control$offsetNodes
    if(is.null(offsetNodes)) stop("Must provide offsetNodes in control list")
    adaptive            <- if(!is.null(control$adaptive))            control$adaptive            else TRUE
    adaptInterval       <- if(!is.null(control$adaptInterval))       control$adaptInterval       else 20 #
    adaptFactorExponent <- if(!is.null(control$adaptFactorExponent)) control$adaptFactorExponent else 0.8
    scale               <- if(!is.null(control$scale))               control$scale               else 1
    ## calculation nodes
    calcNodes <- model$getDependencies(c(target, offsetNodes))
    ## variables for adaptation
    scaleOriginal <- scale
    timesRan      <- 0
    timesAccepted <- 0
    timesAdapted  <- 0
    optimalAR     <- 0.44
    gamma1        <- 0
  },
  run = function() {
    currentTargetValues <- values(model, target)
    currentOffsetNodeValues <- values(model, offsetNodes)
    proposalShift <- rnorm(1, mean = 0, sd = scale)
    proposalTargetValues <- currentTargetValues + proposalShift
    proposalOffsetNodeValues <- currentOffsetNodeValues - proposalShift
    values(model, target) <<- proposalTargetValues
    values(model, offsetNodes) <<- proposalOffsetNodeValues
    logMetropolisHastingsRatio <- calculateDiff(model, calcNodes)
    accept <- decide(logMetropolisHastingsRatio)
    if(accept) nimCopy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
    else     nimCopy(from = mvSaved, to = model, row = 1, nodes = calcNodes, logProb = TRUE)
    if(adaptive)     adaptiveProcedure(accept)
  },
  methods = list(
    adaptiveProcedure = function(accepted = logical()) {
      timesRan <<- timesRan + 1
      if(accepted)     timesAccepted <<- timesAccepted + 1
      if(timesRan %% adaptInterval == 0) {
        acceptanceRate <- timesAccepted / timesRan
        timesAdapted <<- timesAdapted + 1
        gamma1 <<- 1/((timesAdapted + 3)^adaptFactorExponent)
        gamma2 <- 10 * gamma1
        adaptFactor <- exp(gamma2 * (acceptanceRate - optimalAR))
        scale <<- scale * adaptFactor
        timesRan <<- 0
        timesAccepted <<- 0
      }
    },
    reset = function() {
      scale <<- scaleOriginal
      timesRan      <<- 0
      timesAccepted <<- 0
      timesAdapted  <<- 0
      gamma1 <<- 0
    }
  )
)
```

# Let's make an interesting sampler (3)

Configure and build the MCMC.

```{r}
mcmcConf <- configureMCMC(DEmodel)
mcmcConf$addSampler(target = c('sex_int'),
                    type = ourSampler,
                    control = list(offsetNodes = 'farm_effect'))
mcmc <- buildMCMC(mcmcConf)
```

# Accessing more information from a sampler

One might be interested in  accessing compiled internals.  Here is code from an on-the-fly example.

```{r}
## Make compiled model
cDEmodel <- compileNimble(DEmodel)
## Set internal option to access compiled samplers inside of compiled mcmc below
nimbleOptions(buildInterfacesForCompiledNestedNimbleFunctions = TRUE)
## Compile the MCMC
cmcmc <- compileNimble(mcmc)
## See that the run function is an interface to compiled code via .Call
cmcmc$run
## See where our sampler of interest is in the sampler configuration list
mcmcConf$printSamplers()
## And see that the built and compiled samplerFunctions are in the same order
cmcmc$samplerFunctions
## Run the MCMC (could use runMCMC(cmcmc, ...) instead, but time=TRUE is not available via runMCMC)
cmcmc$run(niter = 1000, time = TRUE)
## Access various internals.  Note that some of these have been reset after adaptation steps.
cmcmc$samplerFunctions[[30]]$timesAdapted
cmcmc$samplerFunctions[[30]]$timesAccepted
cmcmc$samplerFunctions[[30]]$scale
cmcmc$samplerFunctions[[30]]$timesRan
## Look at farm_effect in the compiled model
cDEmodel$farm_effect
## Run our sampler 100 times
for(i in 1:100) cmcmc$samplerFunctions[[30]]$run()
## See if there were any updates
cDEmodel$farm_effect
## See how to look at the mvSaved object from the compiled MCMC
## (There are more direct ways to access values in mvSaved. See modelValues
## documentation.)
## This would show the full object: as.matrix(cmcmc$mvSaved)
## This is what I was looking for in the live session:
cmcmc$mvSaved[['farm_effect']]
## See time spent in each sampler (this will not include the 100 iterations
## we did "by hand", only iterations via cmcmc$run(), or runMCMC, which calls
## cmcmc$run).
cmcmc$getTimes()
```


# Let's make an interesting sampler (4)

Run and compare.

```{r, eval=(has_compareMCMCs & generate_orig_results)}
set.seed(123)
DEinits_vals <- DEinits()
mcmcResults_ourSamples <- compareMCMCs(
  modelInfo = list(code = DEcode, 
                   data = list(Ecervi_01 = DeerEcervi$Ecervi_01),
                   constants = DEconstants, # centered
                   inits = DEinits_vals),
  ## monitors ## Use default monitors: top-level parameters
  MCMCs = c('nimble',
            'nimble_custom'),
  nimbleMCMCdefs = list(
    nimble_custom = function(model) {
      mcmcConf <- configureMCMC(model)
      mcmcConf$addSampler(target = c('sex_int'),
                          type = ourSampler,
                          control = list(offsetNodes = 'farm_effect'))
      mcmcConf                         # Output should be an MCMC configuration 
    }
  ),
  MCMCcontrol = list(niter = 20000, burnin = 1000)
)
```
```{r, echo=FALSE, eval=(has_compareMCMCs & generate_orig_results)}
# This is for the original results.
# Run the next one if you want to run it yourself.
make_MCMC_comparison_pages(mcmcResults_ourSamples, modelName = "orig_custom_sampler_results")
```

```{r, eval=FALSE}
# Run this to generate comparison pages on your machine.
make_MCMC_comparison_pages(mcmcResults_ourSamples, modelName = "custom_sampler_results")
```

Results that come with this module are [here](orig_custom_sampler_results.html)

Results if you run it yourself are [here](custom_sampler_results.html)

# Introduction to `modelValues`

- A `modelValues` class does the job of holding multiple sets of values of 
    - model variables and
    - log probabilities ("logProbs") of stochastic nodes.
- We think of each set of values as a "row".
- `modelValues` classes are like `nimbleFunctions` in that:
    - They can be compiled.
    - They should work the same way compiled and uncompiled.
- MCMC output is stored in a `modelValues` object (called `mvSamples`).
- Each `modelValues` class comes from a configuration of what variables are needed (and their types and sizes).
    - Typically the configuration is generated from a model or from variables monitored in MCMC.
    - (Actually, one can hand-configure a `modelValues` in any way one wants.)
- MCMC uses one `modelValues` object configured for **all** model variables and logProbs.
    - Only a single "row" is needed.
- This is the `mvSaved` and must be synchronized with the model at entry and exit of each sampler.

# Recall our use of modelValues in samplers and MCMCs

We used a "1-row" modelValues called `mvSaved` that has the current state of the model.

```{r, eval=FALSE}
if(jump) {
  nimCopy(from = model, to = mvSaved, row = 1, nodes = target, logProb = TRUE)
  nimCopy(from = model, to = mvSaved, row = 1, nodes = calcNodesNoSelfDeterm, logProb = FALSE)
  nimCopy(from = model, to = mvSaved, row = 1, nodes = calcNodesNoSelfStoch, logProbOnly = TRUE)
} else {
  nimCopy(from = mvSaved, to = model, row = 1, nodes = target, logProb = TRUE)
  nimCopy(from = mvSaved, to = model, row = 1, nodes = calcNodesNoSelfDeterm, logProb = FALSE)
  nimCopy(from = mvSaved, to = model, row = 1, nodes = calcNodesNoSelfStoch, logProbOnly = TRUE)
}
```

In the overall MCMC we use a modelValues called `mvSamples` with as many rows as saved iterations that stores the posterior samples for the monitored nodes.

```{r, eval=FALSE}
resize(mvSamples,  mvSamples_copyRow  + floor((niter-nburnin) / thinToUseVec[1]))
nimCopy(from = model, to = mvSamples,  row = mvSamples_copyRow,  nodes = monitors)
```

We'll see more use of modelValues in the next examples on using nimbleFunctions for post-processing.

# nimbleFunctions for post-processing: posterior predictive checks

One standard posterior predictive check is to assess (graphically/informally or formally) whether simulated data from the posterior predictive distribution are similar to the actual data.

We could do this by adding additional 'pseudo-data' nodes to the model, but this will slow down the overall MCMC.

Often we want to do this sort of thing after the MCMC is run, using the samples from the posterior.

Suppose we want to simulate many datasets from the posterior predictive distribution. What do we need to do?

 - Run the MCMC, monitoring all nodes that are parents of the data nodes
 - Iterate through the posterior samples and:
    - Put them in the model
    - Simulate new datasets
    - Store the simulated datasets
 - Carry out our assessment

Note, we could do this fully in R, without using any compiled operations. We could also do this mostly in R but use the compiled model. Either can be fine and using a compiled nimbleFunction may only matter if these other approaches are too slow.

# nimbleFunction for posterior predictive simulation



# later

```
nodes <- cmcmc$mvSamples$getVarNames()
values(m, nodes) <- samples[i, ]
```
