---
title: "Writing new MCMC samplers (or other nimbleFunctions)"
subtitle: "NIMBLE 2021 Virtual Workshop"
author: "NIMBLE Development Team"
date: "May 2021"

output:
  slidy_presentation: default
  beamer_presentation: default
---
<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nimble)
if(!require(coda))
  warning("This Rmd file needs package coda")
has_compareMCMCs <- require(compareMCMCs)
if(!has_compareMCMCs)
  warning("This Rmd file uses package compareMCMCs from github.  Sections using it will be skipped.")
generate_orig_results <- FALSE
```

Agenda
=====

1. Design of NIMBLE's MCMC system
2. Two-stage evaluation of nimbleFunctions with setup code.
3. Setting up a new sampler: A simple Metropolis-Hastings example
4. Worked example: sampler that jointly updates intercepts and random effects in the E. cervi example.

Load Deer E. cervi example
=====
```{r load-DeerEcervi}
source(file.path("..", "examples", "DeerEcervi", "load_DeerEcervi.R"), 
       chdir = TRUE)
set.seed(123)
DEmodel <- nimbleModel(DEcode,
                        constants = DEconstants,
                        data = list(Ecervi_01 = DeerEcervi$Ecervi_01),
                        inits = DEinits())
```


Design of `nimble`'s MCMC system
=====

Here is a figure of MCMC configuration and MCMCs:[nimble_MCMC_design.pdf](nimble_MCMC_design.pdf)

1. MCMC configuration object: Contains a list of sampler assignments, not actual samplers.
2. MCMC object: Contains a list of sampler objects.

To write new samplers, we need to understand:

- Two-stage evaluation of nimbleFunctions with setup code.
- Setup (configuration) rules for using a new MCMC sampler
- Run-time rules for management of model calculations and saved states.
- More about nimbleFunction programming.

Example of MCMC configuration object
=====
Look at MCMC configuration object for just a few nodes
```{r}
mcmcConf <- configureMCMC(DEmodel, nodes = 'farm_effect[1:3]')
class(mcmcConf)
ls(mcmcConf)
fe1_sampler <- mcmcConf$getSamplers("farm_effect[1]")[[1]]
class(fe1_sampler)
ls(fe1_sampler)
fe1_sampler$target
fe1_sampler$control
```

What happens from an MCMC configuration object?
=====

Eventually, an MCMC sampler object is created by a call like this (for adaptive random-walk MH):
```{r, eval=FALSE}
sampler_RW(model, mvSaved, target, control)
```

- This is stage one of two-stage evaluation.  It instantiates an object of a `sampler_RW` class.
- `model` is the model
- `mvSaved` is a `modelValues` object for keeping a set of saved model states.
- `target` is a vector of target node names
- `control` is a list of whatever sampler-specific configuration settings are needed.

Two-stage evaluation in nimbleFunctions
=====
Say we want a nimbleFunction to calculate some nodes and their dependencies.

```{r}
calcDepsNF <- nimbleFunction(
  setup = function(model, nodes) { # setup function gives first stage of evalution
    calcNodes <- model$getDependencies(nodes)
  },
  run = function() {               # run function (or other methods) give sectond stage of evaluation
    ans <- model$calculate(calcNodes)
    return(ans)
    returnType(double())
  }
)
```

Two-stage evaluation in nimbleFunctions
=====
```{r}
calcDeps_farm1 <- calcDepsNF(DEmodel, 'farm_effect[1]') ## first stage: setup code
calcDeps_farm1$run() ## second stage: run code
```

We could compiled `calcDeps_farm1`.  Then `run` would become a method of a C++ class.

Notice that:

- `calcDeps_farm1` is an R reference class object of a custom-generated class.
- `model` and `calcNodes` are fields in the class

```{r}
class(calcDeps_farm1) ## We could have used nimbleFunction's name argument to set the class name
calcDeps_farm1$calcNodes
```

Go to R to demonstrate uncompiled browsing/debugging.
=====
```{r, eval=FALSE}
calcDepsNF <- nimbleFunction(
  setup = function(model, nodes) {
    browser()
    calcNodes <- model$getDependencies(nodes)
  },
  run = function() {
    browser()
    ans <- model$calculate(calcNodes)
    return(ans)
    returnType(double())
  }
) ## warning about not being able to compiled with browser() is expected.
```

Stepping through debugging from `browser()` will not work well in Rmarkdown, so this code is not evaluated.  Run it in your own R session.
```{r, eval=FALSE}
calcDeps_farm1 <- calcDepsNF(DEmodel, 'farm_effect[1]') ## We'll see the setup code followed by internal code.
calcDeps_farm1$run()
```

More about nimbleFunctions
=====

- Without setup code, a `nimbleFunction` becomes an R function (uncompiled) and a C++ function (compiled).
- With setup code, a `nimbleFunction` becomes an R reference class definition (uncompiled) and a C++ class definition (compiled).
    - `nimbleFunction` returns a generator (aka constructor, aka initializer) of new class objects.

### nimbleFunction class definitions (i.e with setup code)

- `setup` is always executed in R.
    - Typically one-time, high-level processing such as querying model structure.
- `run` and other methods can be run uncompiled (in R) or compiled (via C++).
    - Typically repeated "actual algorithm" calculations such as MCMC sampler updates.
    - Can operate models.
- Any objects (e.g. `calcNodes` and `model`) in `setup` can be used in `run`.
    - Internally, these are automatically set up as class member data.
    - You do not need to explicitly declare class member data.
    - Nodes used in model operations are "baked in" (aka partially evaluated) during compilation. 
        - Node vectors must be created in setup code and used in run code.
        - They can't be dynamically modified in run code.

A basic Random-Walk Metropolis-Hastings sampler
=====
```{r}
ourMH <- nimbleFunction(
  name = 'ourMH',                              # Convenient for class name of R reference class and generated C++ class
  contains = sampler_BASE,                     # There is a simple class inheritance system.
  setup = function(model, mvSaved, target, control) {                 # REQUIRED setup arguments
    scale <- if(!is.null(control$scale)) control$scale else 1         # Typical extraction of control choices
    calcNodes <- model$getDependencies(target)                        # Typical query of model structure
  },                                                                  # setup can't return anything
  run = function() {
    currentValue <- model[[target]]                                   # extract current value
    currentLogProb <- model$getLogProb(calcNodes)                     # get log "denominator" from cached values
    proposalValue <- rnorm(1, mean = currentValue, sd = scale)        # generate proposal value
    model[[target]] <<- proposalValue                                 # put proposal value in model
    proposalLogProb <- model$calculate(calcNodes)                     # calculate log "numerator" 
    logAcceptanceRatio <- currentLogProb - proposalLogProb            # log acceptance ratio
    # Alternative:
    # logAcceptanceRatio <- model$calculateDiff(calcNodes)
    accept <- decide(logMHR)                                          # utility function to generate accept/reject decision
    if(accept)                                                        # accept synchronize model -> mvSaved
      copy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
    else                                                              # reject: synchronize mvSaved -> model
      copy(from = mvSaved, to = model, row = 1, nodes = calcNodes, logProb = TRUE)
  },
  methods = list(                              # required method for sampler_BASE base class
    reset = function() {}
  )
)
```

Rules for each sampler
=====

### setup function
  
- The four arguments, named exactly as shown, are required.  This allows `buildMCMC` to create any sampler correctly.

### run function

- The `mvSaved` ("modelValues saved") has a saved copy of all model variables and log probabilities
- Upon entry, `run()` can assume:
    - the model is fully calculated (so `getLogProb` and `calculateDiff` make sense).
    - `mvSaved` and the model are synchronized (have the same values).
- Upon exit, `run()` must ensure those conditions are met.
    - That way the next sampler can operator correctly.
- Between entry and exit, `run()` can manipulate the model in any way necessary.

### reset function

- To match the `sampler_BASE` definition, all samplers must have a `reset()` function.

Introduction to `modelValues`
=====
- A `modelValues` class does the job of holding multiple sets of values of 
    - model variables and
    - log probabilities ("logProbs") of stochastic nodes.
- We think of each set of values as a "row".
- `modelValues` classes are like `nimbleFunctions` in that:
    - They can be compiled.
    - They should work the same way compiled and uncompiled.
- MCMC output is stored in a `modelValues` object (called `mvSamples`).
- Each `modelValues` class comes from a configuration of what variables are needed (and their types and sizes).
    - Typically the configuration is generated from a model or from variables monitored in MCMC.
    - (Actually, one can hand-configure a `modelValues` in any way one wants.)
- MCMC uses one `modelValues` object configured for **all** model variables and logProbs.
    - Only a single "row" is needed.
- This is the `mvSaved` and must be synchronized with the model at entry and exit of each sampler.

Stepping through uncompiled execution
=====

Version with `browser()`s

```{r}
ourMH_debug <- nimbleFunction(
  name = 'ourMH_debug',
  contains = sampler_BASE,
  setup = function(model, mvSaved, target, control) {
    browser()
    scale <- if(!is.null(control$scale)) control$scale else 1 
    calcNodes <- model$getDependencies(target)                
  },                                                          
  run = function() {
    browser()
    currentValue <- model[[target]]                           
    currentLogProb <- model$getLogProb(calcNodes)             
    proposalValue <- rnorm(1, mean = currentValue, sd = scale)
    model[[target]] <<- proposalValue                             
    proposalLogProb <- model$calculate(calcNodes)             
    logAcceptanceRatio <- currentLogProb - proposalLogProb    
    accept <- decide(logMHR)                                     
    if(accept)                                                   
      copy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
    else                                                         
      copy(from = mvSaved, to = model, row = 1, nodes = calcNodes, logProb = TRUE)
  },
  methods = list(                              
    reset = function() {}
  )
)
```

Stepping through uncompiled execution
=====
```{r}
mcmcConf <- configureMCMC(DEmodel, nodes = NULL) ## Make an empty configuration
mcmcConf$addSampler(type = "ourMH_debug", target = 'farm_effect[1]')
```

```{r, eval=FALSE}
# run this on your own to stept through in debug (browser) mode.
mcmc <- buildMCMC(mcmcConf)
mcmc$run(5)
```

Let's make an interesting sampler (1)
=====

We'll put the pieces together for the E. cervi example as follows:

- Use the original parameterization.
- Write a sampler that proposes to add $\delta \sim N(0, \mbox{scale})$ to the two sex-specific intercepts and to subtract the same $\delta$ from every `farm_effect[i]` ($i = 1 \ldots 24$).
- This is a scalar sampler in rotated coordinates.
- The coordinate transformation is linear, so there is no determinant of a Jacobian matrix to incorporate.  In general one needs to be careful to use distribution theory correctly if non-linear coordinate transformations are involved.  We are covering the software, not the math.
- We want proposal scale to be adaptive (self-tuning as the MCMC proceeds).  We will just copy from NIMBLE's `sampler_RW` to implement that.

Let's make an interesting sampler (2)
=====

```{r}
ourSampler <- nimbleFunction(
  name = 'ourSampler',
  contains = sampler_BASE,
  setup = function(model, mvSaved, target, control) {
    ## control list extraction
    offsetNodes <- control$offsetNodes
    if(is.null(offsetNodes)) stop("Must provide offsetNodes in control list")
    adaptive            <- if(!is.null(control$adaptive))            control$adaptive            else TRUE
    adaptInterval       <- if(!is.null(control$adaptInterval))       control$adaptInterval       else 20 #
    adaptFactorExponent <- if(!is.null(control$adaptFactorExponent)) control$adaptFactorExponent else 0.8
    scale               <- if(!is.null(control$scale))               control$scale               else 1
    ## calculation nodes
    calcNodes <- model$getDependencies(c(target, offsetNodes))
    ## variables for adaptation
    scaleOriginal <- scale
    timesRan      <- 0
    timesAccepted <- 0
    timesAdapted  <- 0
    optimalAR     <- 0.44
    gamma1        <- 0
  },
  run = function() {
    currentTargetValues <- values(model, target)
    currentOffsetNodeValues <- values(model, offsetNodes)
    proposalShift <- rnorm(1, mean = 0, sd = scale)
    proposalTargetValues <- currentTargetValues + proposalShift
    proposalOffsetNodeValues <- currentOffsetNodeValues - proposalShift
    values(model, target) <<- proposalTargetValues
    values(model, offsetNodes) <<- proposalOffsetNodeValues
    logMetropolisHastingsRatio <- calculateDiff(model, calcNodes)
    accept <- decide(logMetropolisHastingsRatio)
    if(accept) nimCopy(from = model, to = mvSaved, row = 1, nodes = calcNodes, logProb = TRUE)
    else     nimCopy(from = mvSaved, to = model, row = 1, nodes = calcNodes, logProb = TRUE)
    if(adaptive)     adaptiveProcedure(accept)
  },
  methods = list(
    adaptiveProcedure = function(accepted = logical()) {
      timesRan <<- timesRan + 1
      if(accepted)     timesAccepted <<- timesAccepted + 1
      if(timesRan %% adaptInterval == 0) {
        acceptanceRate <- timesAccepted / timesRan
        timesAdapted <<- timesAdapted + 1
        gamma1 <<- 1/((timesAdapted + 3)^adaptFactorExponent)
        gamma2 <- 10 * gamma1
        adaptFactor <- exp(gamma2 * (acceptanceRate - optimalAR))
        scale <<- scale * adaptFactor
        timesRan <<- 0
        timesAccepted <<- 0
      }
    },
    reset = function() {
      scale <<- scaleOriginal
      timesRan      <<- 0
      timesAccepted <<- 0
      timesAdapted  <<- 0
      gamma1 <<- 0
    }
  )
)
```

Let's make an interesting sampler (3)
=====

Configure and build the MCMC.

```{r}
mcmcConf <- configureMCMC(DEmodel)
mcmcConf$addSampler(target = c('sex_int'),
                    type = ourSampler,
                    control = list(offsetNodes = 'farm_effect'))
mcmc <- buildMCMC(mcmcConf)
```

Live in-session code
=====

Questions arose about accessing compiled internals.  Here is code from an on-the-fly example.

```{r}
## Make compiled model
cDEmodel <- compileNimble(DEmodel)
## Set internal option to access compiled samplers inside of compiled mcmc below
nimbleOptions(buildInterfacesForCompiledNestedNimbleFunctions = TRUE)
## Compile the MCMC
cmcmc <- compileNimble(mcmc)
## See that the run function is an interface to compiled code via .Call
cmcmc$run
## See where our sampler of interest is in the sampler configuration list
mcmcConf$printSamplers()
## And see that the built and compiled samplerFunctions are in the same order
cmcmc$samplerFunctions
## Run the MCMC (could use runMCMC(cmcmc, ...) instead, but time=TRUE is not available via runMCMC)
cmcmc$run(niter = 1000, time = TRUE)
## Access various internals.  Note that some of these have been reset after adaptation steps.
cmcmc$samplerFunctions[[30]]$timesAdapted
cmcmc$samplerFunctions[[30]]$timesAccepted
cmcmc$samplerFunctions[[30]]$scale
cmcmc$samplerFunctions[[30]]$timesRan
## Look at farm_effect in the compiled model
cDEmodel$farm_effect
## Run our sampler 100 times
for(i in 1:100) cmcmc$samplerFunctions[[30]]$run()
## See if there were any updates
cDEmodel$farm_effect
## See how to look at the mvSaved object from the compiled MCMC
## (There are more direct ways to access values in mvSaved. See modelValues
## documentation.)
## This would show the full object: as.matrix(cmcmc$mvSaved)
## This is what I was looking for in the live session:
cmcmc$mvSaved[['farm_effect']]
## See time spent in each sampler (this will not include the 100 iterations
## we did "by hand", only iterations via cmcmc$run(), or runMCMC, which calls
## cmcmc$run).
cmcmc$getTimes()
```


Let's make an interesting sampler (4)
=====

Run and compare.

```{r, eval=(has_compareMCMCs & generate_orig_results)}
set.seed(123)
DEinits_vals <- DEinits()
mcmcResults_ourSamples <- compareMCMCs(
  modelInfo = list(code = DEcode, 
                   data = list(Ecervi_01 = DeerEcervi$Ecervi_01),
                   constants = DEconstants, # centered
                   inits = DEinits_vals),
  ## monitors ## Use default monitors: top-level parameters
  MCMCs = c('nimble',
            'nimble_custom'),
  nimbleMCMCdefs = list(
    nimble_custom = function(model) {
      mcmcConf <- configureMCMC(model)
      mcmcConf$addSampler(target = c('sex_int'),
                          type = ourSampler,
                          control = list(offsetNodes = 'farm_effect'))
      mcmcConf                         # Output should be an MCMC configuration 
    }
  ),
  MCMCcontrol = list(niter = 20000, burnin = 1000)
)
```
```{r, echo=FALSE, eval=(has_compareMCMCs & generate_orig_results)}
# This is for the original results.
# Run the next one if you want to run it yourself.
make_MCMC_comparison_pages(mcmcResults_ourSamples, modelName = "orig_custom_sampler_results")
```

```{r, eval=FALSE}
# Run this to generate comparison pages on your machine.
make_MCMC_comparison_pages(mcmcResults_ourSamples, modelName = "custom_sampler_results")
```


Results that come with this module are [here](orig_custom_sampler_results.html)

Results if you run it yourself are [here](custom_sampler_results.html)
